---
title: "Twitter Sentiment Classifier"
subtitle: "PSTAT 131"
author: "Aaron Chen-Xiao, Yumeng Guo, Sophia Leng, Sahand Shafieha, Olivia Shao"
date: "06/12/2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: false
    code_folding: hide
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = T,
                      results = 'markup',
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(data.table)
library(dplyr)
library(tidymodels)
library(lubridate)
library(hms)
library(stringr)
library(tidytext)
library(wordcloud)
library(igraph)
library(ggraph)
```

## Data Cleaning

Here is what was done in the data cleaning process:

1.  Added column names per project_memo submitted previously
2.  Checked for the following:
    1.  Missing values
    2.  Duplicated rows
3.  Checked the levels of columns and removed the columns that had one level and weren't useful
4.  Split the `date` column into `date` and `time_pdt`
5.  Binarized sentiment
6.  Cleaned up `text` column
7.  Tokenized and removed stop words for both unigrams and bigrams for future feature extraction

```{r}
# load the dataset
twitter <- read_csv("sample_50k.csv")
# adding back the column names
colnames(twitter) <- c("sentiment", "id", "date", "query", "user", "text")

# Balanced Dataset
twitter %>%
  count(sentiment)
# Check for missing values
colSums(is.na(twitter))
# Check summary stats for sentiment
summary(twitter$sentiment)
# Check for duplicated rows
sum(duplicated(twitter))

# View levels of query
unique(twitter$query) 
# We only have 1 level "NO_QUERY" so we can drop it
twitter_cleaned <- twitter %>% select(-query)

# Split date column into date, time, and time_zone
twitter_cleaned <- twitter_cleaned %>%
  mutate(date_raw = date) %>%
  mutate(
    # Extract timezone before parsing
    time_zone = sub(".*\\s(\\S+)\\s\\d{4}$", "\\1", date_raw),
    
    # Strip timezone from string, then parse
    parsed_dt = parse_date_time(
      sub("(.*?)\\s+\\S+(\\s+\\d{4})$", "\\1\\2", date_raw),
      orders = "a b d H:M:S Y"
    ),
    
    # Extract and format date as MM/DD/YYYY, stored as Date type
    date = as.Date(format(parsed_dt, "%m/%d/%Y"), format = "%m/%d/%Y"),
    
    # Extract time and store as hms
    time = as_hms(format(parsed_dt, "%H:%M:%S"))
  ) %>%
  select(-parsed_dt, -date_raw)

unique(twitter_cleaned$time_zone)
# I noticed that every time zone in the dataset is in PDT so I drop the column.
# And renamed the time column into time_pdt
twitter_cleaned <- twitter_cleaned %>% select(-time_zone)
twitter_cleaned <- twitter_cleaned %>%
  rename(time_pdt = time)

# Binarize sentiment (0 = negative, 1 = positive)
twitter_cleaned <- twitter_cleaned %>%
  mutate(sentiment = factor(sentiment, levels = c(0, 1), labels = c("Negative", "Positive")))

# Clean up the text
twitter_cleaned <- twitter_cleaned %>%
  mutate(text = str_to_lower(text),                          # lowercase
         text = str_remove_all(text, "@\\w+"),               # remove @mentions
         text = str_remove_all(text, "http\\S+|www\\S+"),    # remove URLs
         text = str_remove_all(text, "\\bi['’]ll\\b"),       # remove contractions
         text = str_remove_all(text, "\\bi['’]d\\b"),
         text = str_remove_all(text, "\\bi['’]m\\b"),
         text = str_remove_all(text, "\\bi['’]ve\\b"),
         text = str_remove_all(text, "[^a-z\\s]"),           # remove punctuation/numbers
         text = str_squish(text))                            # remove extra whitespace

```


```{r}
# First create row_id entry for each tweet, select only row_id, sentiment vaue, and text
twitter_cleaned <- twitter_cleaned %>%
  mutate(row_id = row_number()) %>%
  select(row_id, sentiment, text)
```

### Testing/Training Split
We split our data at this point to avoid data leakage from our TF-IDF, which allows us to get a more accurate estimate of our model performance. 

```{r}
# Testing Training Split
set.seed(1)

twitter_split <- initial_split(
  twitter_cleaned,
  prop = 0.80,
  strata = sentiment
)

twitter_train <- training(twitter_split)
twitter_test <- testing(twitter_split)
```

## Tokenizing/TF-IDF for Training Data

```{r}
# Unigrams TF-IDF
unigram_train <- twitter_train %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  count(row_id, sentiment, word, name = "n") %>%
  bind_tf_idf(
    term = word,
    document = row_id,
    n = n
  )

head(unigram_train, n = 10)

top_unigram_words <- unigram_train %>%
  count(word, sort = TRUE) %>%
  slice_head(n = 5000) %>%
  pull(word)

unigram_train <- unigram_train %>%
  filter(word %in% top_unigram_words)

unigram_train_wide <- unigram_train %>%
  select(row_id, sentiment, word, tf_idf) %>%
  pivot_wider(
    names_from = word,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "word_"
  )
```

```{r}
# Bigrams TF-IDF
bigram_train <- twitter_train %>%
  unnest_tokens(bigram, text, token = "ngrams", n=2)  %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(row_id, sentiment, bigram, name = "n") %>%
  bind_tf_idf(
    term = bigram,
    document = row_id,
    n = n
  )

head(bigram_train, n = 10)

top_bigram_words <- bigram_train %>%
  count(bigram, sort = TRUE) %>%
  slice_head(n = 5000) %>%
  pull(bigram)

bigram_train <- bigram_train %>%
  filter(bigram %in% top_bigram_words)

bigram_train_wide <- bigram_train %>%
  select(row_id, sentiment, bigram, tf_idf) %>%
  pivot_wider(
    names_from = bigram,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "bigram_"
  )
```

## Visualizations

### Frequency Word Cloud
```{r}
unigram_word_counts <- unigram_train %>%
  group_by(word) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n))

wordcloud(
  words = unigram_word_counts$word,
  freq = unigram_word_counts$total_n,
  max.words = 100,
  random.order = FALSE,
  colors = brewer.pal(8, "Dark2")
)
```


### Bigram Network Graph

Note swine flu, jonas brothers, harry potter - dataset is from 2009
```{r}

bigram_train_sep <- twitter_cleaned %>%
  unnest_tokens(bigram, text, token = "ngrams", n=2)  %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  drop_na(word1, word2) %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) 

bigram_graph <- bigram_train_sep %>%
  count(word1, word2) %>%
  filter(n > 20) %>%
  graph_from_data_frame()

a <- grid::arrow(type = "closed", length = unit(.15, "inches"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE,
                 arrow = a, end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void()
```

