---
title: "Twitter Sentiment Classifier"
subtitle: "PSTAT 131"
author: "Aaron Chen-Xiao, Yumeng Guo, Sophia Leng, Sahand Shafieha, Olivia Shao"
date: "06/12/2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: false
    code_folding: hide
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = T,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(data.table)
library(dplyr)
library(tidymodels)
library(lubridate)
library(hms)
library(stringr)
library(tidytext)
```

Here is what was done in the data cleaning process:

1.  Added column names per project_memo submitted previously
2.  Checked for the following:
    1.  Missing values
    2.  Duplicated rows
3.  Checked the levels of columns and removed the columns that had one level and weren't useful
4.  Split the `date` column into `date` and `time_pdt`
5.  Binarized sentiment
6.  Cleaned up `text` column
7.  Tokenized and removed stop words for both unigrams and bigrams for future feature extraction

```{r}
# load the dataset
twitter <- read_csv("sample_50k.csv")
# adding back the column names
colnames(twitter) <- c("sentiment", "id", "date", "query", "user", "text")

# Check for missing values
colSums(is.na(twitter))
# Check summary stats for sentiment
summary(twitter$sentiment)
# Check for duplicated rows
sum(duplicated(twitter))

# View levels of query
unique(twitter$query) 
# We only have 1 level "NO_QUERY" so we can drop it
twitter_cleaned <- twitter %>% select(-query)

# Split date column into date, time, and time_zone
twitter_cleaned <- twitter_cleaned %>%
  mutate(date_raw = date) %>%
  mutate(
    # Extract timezone before parsing
    time_zone = sub(".*\\s(\\S+)\\s\\d{4}$", "\\1", date_raw),
    
    # Strip timezone from string, then parse
    parsed_dt = parse_date_time(
      sub("(.*?)\\s+\\S+(\\s+\\d{4})$", "\\1\\2", date_raw),
      orders = "a b d H:M:S Y"
    ),
    
    # Extract and format date as MM/DD/YYYY, stored as Date type
    date = as.Date(format(parsed_dt, "%m/%d/%Y"), format = "%m/%d/%Y"),
    
    # Extract time and store as hms
    time = as_hms(format(parsed_dt, "%H:%M:%S"))
  ) %>%
  select(-parsed_dt, -date_raw)

unique(twitter_cleaned$time_zone)
# I noticed that every time zone in the dataset is in PDT so I drop the column.
# And renamed the time column into time_pdt
twitter_cleaned <- twitter_cleaned %>% select(-time_zone)
twitter_cleaned <- twitter_cleaned %>%
  rename(time_pdt = time)

# Binarize sentiment (0 = negative, 1 = positive)
twitter_cleaned <- twitter_cleaned %>%
  mutate(sentiment = factor(sentiment, levels = c(0, 1), labels = c("Negative", "Positive")))

# Clean up the text
twitter_cleaned <- twitter_cleaned %>%
  mutate(text = str_to_lower(text),                          # lowercase
         text = str_remove_all(text, "@\\w+"),               # remove @mentions
         text = str_remove_all(text, "http\\S+|www\\S+"),    # remove URLs
         text = str_remove_all(text, "[^a-z\\s]"),           # remove punctuation/numbers
         text = str_squish(text))                            # remove extra whitespace

# Create unigrams (single words)
twitter_unigrams <- twitter_cleaned %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word")

# Create bigrams (two words)
twitter_bigrams <- twitter_cleaned %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  # Remove bigrams that contain stop words in either word
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word) %>%
  unite(bigram, word1, word2, sep = " ")

# View samples
head(twitter_unigrams, n = 10)
head(twitter_bigrams, n = 10)
```

```{=html}
<script>
window.onload = function() {
  var buttons = document.querySelectorAll("button.code-folding-btn");
  buttons.forEach(function(btn) {
    if (btn.innerText.trim() === "Show") {
      btn.innerText = "Show Code";
    }
    if (btn.innerText.trim() === "Hide") {
      btn.innerText = "Hide Code";
    }
  });
};
</script>
```
