---
title: "Twitter Sentiment Classifier"
subtitle: "PSTAT 131"
author: "Aaron Chen-Xiao, Yumeng Guo, Sophia Leng, Sahand Shafieha, Olivia Shao"
date: "06/12/2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: false
    code_folding: hide
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = T,
                      results = 'markup',
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(data.table)
library(dplyr)
library(tidymodels)
library(lubridate)
library(hms)
library(stringr)
library(tidytext)
library(wordcloud)
library(igraph)
library(ggraph)
```

## Data Cleaning

Here is what was done in the data cleaning process:

1.  Added column names per project_memo submitted previously
2.  Checked for the following:
    1.  Missing values
    2.  Duplicated rows
3.  Checked the levels of columns and removed the columns that had one level and weren't useful
4.  Split the `date` column into `date` and `time_pdt`
5.  Binarized sentiment
6.  Cleaned up `text` column
7.  Tokenized and removed stop words for both unigrams and bigrams for future feature extraction

```{r}
# load the dataset
twitter <- read_csv("sample_50k.csv")
# adding back the column names
colnames(twitter) <- c("sentiment", "id", "date", "query", "user", "text")

# Balanced Dataset
twitter %>%
  count(sentiment)
# Check for missing values
colSums(is.na(twitter))
# Check summary stats for sentiment
summary(twitter$sentiment)
# Check for duplicated rows
sum(duplicated(twitter))

# View levels of query
unique(twitter$query) 
# We only have 1 level "NO_QUERY" so we can drop it
twitter_cleaned <- twitter %>% select(-query)

# Split date column into date, time, and time_zone
twitter_cleaned <- twitter_cleaned %>%
  mutate(date_raw = date) %>%
  mutate(
    # Extract timezone before parsing
    time_zone = sub(".*\\s(\\S+)\\s\\d{4}$", "\\1", date_raw),
    
    # Strip timezone from string, then parse
    parsed_dt = parse_date_time(
      sub("(.*?)\\s+\\S+(\\s+\\d{4})$", "\\1\\2", date_raw),
      orders = "a b d H:M:S Y"
    ),
    
    # Extract and format date as MM/DD/YYYY, stored as Date type
    date = as.Date(format(parsed_dt, "%m/%d/%Y"), format = "%m/%d/%Y"),
    
    # Extract time and store as hms
    time = as_hms(format(parsed_dt, "%H:%M:%S"))
  ) %>%
  select(-parsed_dt, -date_raw)

unique(twitter_cleaned$time_zone)
# I noticed that every time zone in the dataset is in PDT so I drop the column.
# And renamed the time column into time_pdt
twitter_cleaned <- twitter_cleaned %>% select(-time_zone)
twitter_cleaned <- twitter_cleaned %>%
  rename(time_pdt = time)

# Binarize sentiment (0 = negative, 1 = positive)
twitter_cleaned <- twitter_cleaned %>%
  mutate(sentiment = factor(sentiment, levels = c(0, 1), labels = c("Negative", "Positive")))

# Clean up the text
twitter_cleaned <- twitter_cleaned %>%
  mutate(text = str_to_lower(text),                          # lowercase
         text = str_remove_all(text, "@\\w+"),               # remove @mentions
         text = str_remove_all(text, "http\\S+|www\\S+"),    # remove URLs
         text = str_remove_all(text, "\\bi['’]ll\\b"),       # remove contractions
         text = str_remove_all(text, "\\bi['’]d\\b"),
         text = str_remove_all(text, "\\bi['’]m\\b"),
         text = str_remove_all(text, "\\bi['’]ve\\b"),
         text = str_remove_all(text, "[^a-z\\s]"),           # remove punctuation/numbers
         text = str_squish(text))                            # remove extra whitespace

```


```{r}
# First create row_id entry for each tweet, select only row_id, sentiment vaue, and text
twitter_cleaned <- twitter_cleaned %>%
  mutate(row_id = row_number()) %>%
  select(row_id, sentiment, text)
```

## Testing/Training Split
We split our data at this point to avoid data leakage from our TF-IDF, which allows us to get a more accurate estimate of our model performance. 

```{r}
# Testing Training Split
set.seed(1)

twitter_split <- initial_split(
  twitter_cleaned,
  prop = 0.80,
  strata = sentiment
)

twitter_train <- training(twitter_split)
twitter_test <- testing(twitter_split)
```

## Tokenizing/TF-IDF

### Training Data

```{r}
# Unigrams TF-IDF
unigram_train <- twitter_train %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  count(row_id, sentiment, word, name = "n") %>%
  bind_tf_idf(
    term = word,
    document = row_id,
    n = n
  )

head(unigram_train, n = 10)

top_unigram_words <- unigram_train %>%
  count(word, sort = TRUE) %>%
  slice_head(n = 5000) %>%
  pull(word)

unigram_train <- unigram_train %>%
  filter(word %in% top_unigram_words)

unigram_train_wide <- unigram_train %>%
  select(row_id, sentiment, word, tf_idf) %>%
  pivot_wider(
    names_from = word,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "word_"
  )
```


```{r}
# Bigrams TF-IDF
bigram_train <- twitter_train %>%
  unnest_tokens(bigram, text, token = "ngrams", n=2)  %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(row_id, sentiment, bigram, name = "n") %>%
  bind_tf_idf(
    term = bigram,
    document = row_id,
    n = n
  )

head(bigram_train, n = 10)

top_bigram_words <- bigram_train %>%
  count(bigram, sort = TRUE) %>%
  slice_head(n = 5000) %>%
  pull(bigram)

bigram_train <- bigram_train %>%
  filter(bigram %in% top_bigram_words)

bigram_train_wide <- bigram_train %>%
  select(row_id, sentiment, bigram, tf_idf) %>%
  pivot_wider(
    names_from = bigram,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "bigram_"
  )
```

### Testing Data

We have to put the testing data in the same format. However, we don't want the TF-IDF values for words that were only in the test set, because if we are treating it as 'new data' we wouldn't have access to it.

```{r}
# Unigram Testing Data
unigram_test <- twitter_test %>%
  select(row_id, sentiment, text) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  filter(word %in% top_unigram_words) %>%  # keep only words that were used in train tf-idf
  count(row_id, sentiment, word, name = "n") %>%
  bind_tf_idf(
    term = word,
    document = row_id,
    n = n
  )

unigram_test_wide <- unigram_test %>%
  select(row_id, sentiment, word, tf_idf) %>%
  pivot_wider(
    names_from = word,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "word_"
  )

# to match number of cols, fill in 0's for words that were in training but not in testing
# after, both data frames should have 5002 columns (5000 words + sentiment + row_id)
missing_unigram_cols <- setdiff(
  names(unigram_train_wide),
  names(unigram_test_wide)
)

unigram_test_wide[missing_unigram_cols] <- 0

unigram_test_wide <- unigram_test_wide %>%
  select(all_of(names(unigram_train_wide)))
```

```{r}
# Bigram Testing data
bigram_test <- twitter_test %>%
  select(row_id, sentiment, text) %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, into = c("word1", "word2"), sep = " ") %>%
  drop_na(word1, word2) %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%
  unite(bigram, word1, word2, sep = " ") %>%
  filter(bigram %in% top_bigram_words) %>%   #do same as unigrams
  count(row_id, sentiment, bigram, name = "n") %>%
  bind_tf_idf(
    term = bigram,
    document = row_id,
    n = n
  )

bigram_test_wide <- bigram_test %>%
  select(row_id, sentiment, bigram, tf_idf) %>%
  pivot_wider(
    names_from = bigram,
    values_from = tf_idf,
    values_fill = 0,
    names_prefix = "bigram_"
  )

# Match column number again
missing_bigram_cols <- setdiff(
  names(bigram_train_wide),
  names(bigram_test_wide)
)

bigram_test_wide[missing_bigram_cols] <- 0

bigram_test_wide <- bigram_test_wide %>%
  select(all_of(names(bigram_train_wide)))
```


## Visualizations

### Frequency Word Cloud
```{r}
unigram_word_counts <- unigram_train %>%
  group_by(word) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n))

wordcloud(
  words = unigram_word_counts$word,
  freq = unigram_word_counts$total_n,
  max.words = 100,
  random.order = FALSE,
  colors = brewer.pal(8, "Dark2")
)
```


### Bigram Network Graph

Note swine flu, jonas brothers, harry potter - dataset is from 2009
```{r}

bigram_train_sep <- twitter_cleaned %>%
  unnest_tokens(bigram, text, token = "ngrams", n=2)  %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  drop_na(word1, word2) %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) 

bigram_graph <- bigram_train_sep %>%
  count(word1, word2) %>%
  filter(n > 20) %>%
  graph_from_data_frame()

a <- grid::arrow(type = "closed", length = unit(.15, "inches"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE,
                 arrow = a, end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void()
```










## Sparsity of Test Matrices

This is useful for the results later -  the first code chunk shows how sparse the test dataframe is, the second shows the median number of words that was matched to TF-IDF values from the training. This is important because the more words that are the same, the better our model is going to do.

### Unigram Sparsity
```{r}
unigram_predictors <- unigram_test_wide %>%
  select(-row_id, -sentiment)

mean(unigram_predictors == 0)
```

```{r}
unigram_test_wide %>%
  mutate(nonzero_unigrams = rowSums(across(-c(row_id, sentiment), ~ .x != 0))) %>%
  summarise(
    min_nonzero = min(nonzero_unigrams),
    mean_nonzero = mean(nonzero_unigrams),
    median_nonzero = median(nonzero_unigrams),
    max_nonzero = max(nonzero_unigrams)
  )
```


### Bigram Sparsity

Note that for bigrams the median is lower so the model is likely going to do worse. This is partly because tweets are very short, so each tweet only has a few bigrams and since bigrams are more unique we are less likely to have seen it in the training data.

```{r}
bigram_predictors <- bigram_test_wide %>%
  select(-row_id, -sentiment)

mean(bigram_predictors == 0)
```

```{r}
bigram_test_wide %>%
  mutate(nonzero_bigrams = rowSums(across(-c(row_id, sentiment), ~ .x != 0))) %>%
  summarise(
    min_nonzero = min(nonzero_bigrams),
    mean_nonzero = mean(nonzero_bigrams),
    median_nonzero = median(nonzero_bigrams),
    max_nonzero = max(nonzero_bigrams)
  )
```

